---
title: "White Wine Exploratory Data Analysis"
author: "Lekhraj Sharma"
date: "Sept 2015"
output:
  html_document:
    keep_md: no
    theme: readable
    toc: yes
    number_sections: true
    fig_caption: true
---

# About

This exploratory data analysis is towards **"Project 4: Explore and Summarize Data"**, a part of **Data Analyst Nanodegree from Udacity**.

In this exercise, we will explore a data set on white wine quality and physicochemical properties.  The objective is to explore which chemical properties influence the quality of white wines. 

## Dataset  

Dataset chosen is "White Wine Quality" tidy data set downloaded from [Udacity site](https://s3.amazonaws.com/udacity-hosted-downloads/ud651/wineQualityWhites.csv). This publicly available data set was created by Paulo Cortez, etc in 2009 for white variants of the Portuguese "Vinho Verde" wine. Due to privacy and logistic issues, only physicochemical (inputs) and sensory (the output) variables are available (e.g. there is no data about grape types, wine brand, wine selling price, etc.).

This tidy data set contains 4,898 white wines with 11 variables on quantifying the chemical properties of each wine. At least 3 wine experts rated the quality of each wine, providing a rating between 0 (very bad) and 10 (very excellent).

More information about this data set is available at [Udacity site](https://s3.amazonaws.com/udacity-hosted-downloads/ud651/wineQualityInfo.txt).


### Variable description  

Description of attributes:  

1. **Fixed acidity** (tartaric acid - g / dm^3) : most acids involved with wine or fixed or nonvolatile (do not evaporate readily)  
2. **Volatile acidity** (acetic acid - g / dm^3) : the amount of acetic acid in wine, which at too high of levels can lead to an unpleasant, vinegar taste  
3. **Citric acid** (g / dm^3) : found in small quantities, citric acid can add 'freshness' and flavor to wines  
4. **Residual sugar** (g / dm^3) : the amount of sugar remaining after fermentation stops, it's rare to find wines with less than 1 gram/liter and wines with greater than 45 grams/liter are considered sweet  
5. **Chlorides** (sodium chloride - g / dm^3): the amount of salt in the wine  
6. **Free sulfur dioxide** (mg / dm^3): the free form of SO2 exists in equilibrium between molecular SO2 (as a dissolved gas) and bisulfite ion; it prevents microbial growth and the oxidation of wine  
7. **Total sulfur dioxide** (mg / dm^3): amount of free and bound forms of S02; in low concentrations, SO2 is mostly undetectable in wine, but at free SO2 concentrations over 50 ppm, SO2 becomes evident in the nose and taste of wine  
8. **Density** (g / cm^3): the density of water is close to that of water depending on the percent alcohol and sugar content 
9. **pH**: describes how acidic or basic a wine is on a scale from 0 (very acidic) to 14 (very basic); most wines are between 3-4 on the pH scale  
10. **Sulphates** (potassium sulphate - g / dm3): a wine additive which can contribute to sulfur dioxide gas (S02) levels, which acts as an antimicrobial and antioxidant  
11. **Alcohol** (% by volume): The percent alcohol content of the wine  
12. **Quality** (score between 0 to 10) : output variable based on sensory data, rating between 0 (very bad) and 10 (very excellent)
   
## Question  

In this data exploration we wish to answer following question:  

1. *Which chemical properties influence the quality of white wines?*

---

# Load Relevant Libraries

```{r, echo=FALSE, message=FALSE, warning=FALSE, packages}
library(ggplot2)
library(psych)
library(memisc)
library(scatterplot3d)
library(gridExtra)
```

# Load the Data 

Before we can start our data exploration, we need to load the data-set for white wines in a data frame.

```{r echo=FALSE, Load_the_Data}
wine <- read.csv("./wineQualityWhites.csv")
str(wine)
```

# Summary of the dataset

Let us start with getting basic descriptive statistics about the data-set. We will pay extra attention to **quality** as we are trying to answer question on how can we determine wine quality.


```{r echo=FALSE}
summary(wine)
describe(wine, fast=T )

```

Both `summary` and `describe` commands give us interesting insights about our attributes. Clearly attribute **X** is just a serial id for each observation. All other attributes except **quality** have a continuous value with their mean and median being close, therefore the values are distributed evenly across the total range. **Quality** has discrete values ranging from 3 to 9. 

# Univariate Plots and Analysis

Let us create boxplot for our attributes (leaving **X** out as it is a serial id) to see visually how they are distributed.

```{r echo=FALSE, Univariate_Plots}
par(mfrow=c(3,4)) # Put 4 plots in a row
for (i in 2:length(wine)) { # We skip first one as "X" is just a serial id
     boxplot(wine[i], main=names(wine[i])) }
```

Boxplot is confirming what we observed in `summary` that mean and median are close for our attributes here. We can see that other than **alcohol** most other attributes have outliers. We also see that most of the **quality** values lie between 4 to 7.

Let us continue our exploration of seeing how our attributes are distributed by looking at their histograms. 

```{r echo=FALSE}
par(mfrow=c(3,4)) # Put 4 plots in a row
for (i in 2:length(wine)) { # We skip first one as "X" is just a serial id
     hist(wine[,i], xlab=names(wine[i]), main=NULL) }
```

We can see from histograms that almost all attributes other than **residual.sugar** and **alcohol** seem to have normal distribution. Both **residual.sugar** and **alcohol** are left skewed (long tail on right) and can be converted to normal distribution with some transformations. There is further confirmation of our earlier observation that values are distributed evenly across the total range and **quality** has discrete integer values.

## Wine quality    

Let us look into distribution of quality deeper.

**Overall Summary**
```{r echo=FALSE}
summary(wine$quality)
```

**IQR**
```{r echo=FALSE}
IQR(wine$quality)
```

**Frequency Table**
```{r echo=FALSE}
table(wine$quality)
```

So **quality** discrete values range between 3 and 9 with no scores below 3 or above 9. As we saw in earlier boxplot we do have outliers in quality, which is confirmed by seeing that both max and min are more than 1.5*IQR away from their respective quartile. From `table` we can see we have around 4% (200/4898) outliers which is significant. 

Let us plot quality histogram to visualize this better. 

```{r echo=FALSE}
# Utility function to plot a univariate histogram plot for given variable of
# interest and its bindwidth
ggCustomHistogram <- function(hist_var, hist_binwidth) { 
  return(
    ggplot(data=wine, aes_string(x=hist_var)) + 
    geom_bar(binwidth=hist_binwidth, color='black', fill='green') +
    geom_vline(xintercept = mean(wine[,hist_var]), linetype= 'longdash',
               color='red', alpha=.5) +               
    geom_vline(xintercept = median(wine[,hist_var]), linetype='longdash',
               color='blue', alpha=.5) + 
    geom_vline(
      xintercept = quantile(wine[,hist_var], 0.25) - 1.5 * IQR(wine[,hist_var]),
      linetype='dashed', alpha=.5) +
    geom_vline(
      xintercept = quantile(wine[,hist_var], 0.75) + 1.5 * IQR(wine[,hist_var]),
      linetype='dashed', alpha=.5) )
}

ggCustomHistogram("quality", 1)
```

*We can see that majority of wines have quality scores ranging from 5 to 7. Therefore there are lot more average quality wines and few poor or excellent quality wines.* 

Let us explore few more variables such as alcohol, pH and various acids as they are typically mentioned in literature related to wine quality.

## Alcohol   

Let us first look into distribution of alcohol deeper.

**Overall Summary**
```{r echo=FALSE}
summary(wine$alcohol)
```

**IQR**
```{r echo=FALSE}
IQR(wine$alcohol)
```

So **alcohol** numeric values range between 8 and 14.5 with mean at 10.4.

Let us plot alcohol histogram to visualize this better. 

```{r echo=FALSE}
ggCustomHistogram("alcohol", 0.5)
```

As we saw in earlier boxplot we do **not** have outliers in alcohol, which is confirmed by seeing that both max and min are within the 1.5*IQR range from their respective quartile. We have almost normal distribution exceptional peak at 9. As median and mean are quite close, we have reasonably even distribution around mean.

## Citric.Acid   

Let us look into distribution of citric.acid deeper.

**Overall Summary**
```{r echo=FALSE}
summary(wine$citric.acid)
```

**IQR**
```{r echo=FALSE}
IQR(wine$citric.acid)
```

So **citric.acid** numeric values range between 0.0 and 1.66 with mean at 0.33.

Let us plot citric.acid histogram to visualize this better. 

```{r echo=FALSE}
ggCustomHistogram("citric.acid", 0.05)
```

As we saw in earlier boxplot we do have few outliers in citric.acid, which is confirmed by seeing that both min and max are more than 1.5*IQR range away from their respective quartile. We have decent normal distribution. As median and mean are quite close, we have reasonably even distribution around mean.


## Volatile.Acidity   

Let us look into distribution of volatile.acidity deeper.

**Overall Summary**
```{r echo=FALSE}
summary(wine$volatile.acidity)
```

**IQR**
```{r echo=FALSE}
IQR(wine$volatile.acidity)
```

So **volatile.acidity** numeric values range between 0.08 and 1.1 with mean at 0.278.

Let us plot volatile.acidity histogram to visualize this better. 

```{r echo=FALSE}
ggCustomHistogram("volatile.acidity", 0.05)
```

As we saw in earlier boxplot we do have few outliers in volatile.acidity, which is confirmed by seeing that max is than 1.5*IQR range away from 3rd quartile. We have decent normal distribution with slight tail on right. As median and mean are quite close, we have reasonably even distribution around mean.

## pH   

Let us look into distribution of pH deeper.

**Overall Summary**
```{r echo=FALSE}
summary(wine$pH)
```

**IQR**
```{r echo=FALSE}
IQR(wine$pH)
```

So **pH** numeric values range between 2.72 and 3.82 with mean at 3.188.

Let us plot alcohol histogram to visualize this better. 

```{r echo=FALSE}
ggCustomHistogram("pH", 0.05)
```

As we saw in earlier boxplot we do have few outliers in pH, which is confirmed by seeing that both max and min are more than 1.5*IQR range away from their respective quartile. We have good normal distribution. As median and mean are quite close, we have reasonably even distribution around mean.


## Short questions

### What is the structure of your dataset?
We have a list of 4898 unnamed white wines.  They have all been rated on quality and 13 variables of each wine have been given.  Each value given is numeric other than quality and X which are integers.  The quality rating scale is from 1-10 but this dataset only has value from 3-9. A majority of the wines are in middle of the pack (average wines) with quality score of 5 and 6.

### What is/are the main feature(s) of interest in your dataset?
The main feature of interest is the output feature **quality**. We hope to find out what input features can most accurately predict the quality of a white wine.

### What other features in the dataset do you think will help support your investigation into your feature(s) of interest?
Based on how we see wines listed and cataloged, it would be good to include **alcohol** and **pH** in our model. Other features which impact the taste such as variety of **acid** related feature may also impact the quality of wine.        

### Did you create any new variables from existing variables in the dataset?
Not so far. But we do it later by creating a new variable to classify wines into **bad**, **average** and **good** quality wines.

### Of the features you investigated, were there any unusual distributions? Did you perform any operations on the data to tidy, adjust, or change the form of the data? If so, why did you do this?
We can see that other than **alcohol** most other attributes have outliers. We also see that most of the **quality** values lie between 4 to 7.  Did **not** make any major adjustments to the structure of the data.   

---

# Bivariate Plots and Analysis

Let us start zooming into potential variables that may influence quality of white wine. Starting point would be  to explore correlation between our variables.

## Finding attributes which influence quality  

Before we zoom into the set of attributes which impact the **quality** most, let us create pairwise plot for each of the attributes to see in one glance how do not only each of these attributes are distributed but get top level of idea of how they may be related to each other.  


```{r echo=FALSE, Bivariate_Plots}
# Part of psych package
pairs.panels(wine[2:13], show.points=F, rug=F, density=F, gap=0, ellipses=F, cor=F)
```

Our pairwise plot via `pairs.panel` gives us visual birds eyeview into how each variable is distributed along with how they relate to each other. We can see that there are some strong correlation among other attributes such as **0.84** between **density** and **residual.sugar**, which can be seen in pairwise graph as well in terms of almost 45 degree line between these two attributes. Since here we are focusing on **quality** relation with other attributes, we are **not** going to explore these relationships by themselves other than in context of attributes that relate to **quality**.

Let us look deeper into these correlations further with a heatmap plot.

```{r echo=FALSE}
options(width = 100)
# These are part of psych library
lowerCor(wine[2:13])
cor.plot(cor(wine[2:13]))
```

Our correlation plot provides a nice visual view with darker shaded color indicating stronger correlation. With this visual cue and looking in pairwise correlation matrix, we can see that **quality** has strong correlation with **alcohol** (0.44), **density**(-0.31), **chlorides** (-0.21), **volatile.acidity** (-0.19), **toal.sulfur.dioxide** (-0.17) and bit weaker correlation with **fixed.acidity** (-0.11), **residual.sugar** (-0.10), **pH** (0.10).

When we look further in pairwise correlation plot and matrix, we can see that several of the strongly correlated attributes to **quality** have good correlation among themselves. Since we want to identify independent variables which impact the quality, it will be good to prune these out and select one with strongest correlation with **quality**. We see that **density** (-0.78), **total.sulphur.dioxide** (-0.45), **chlorides** (-0.36) and **residual.sugar** (-0.45) all have good correlation with our strongly correlated attributed **alcohol**. Similarly **pH** and **fixed.acidity** also have strong correlation (-0.43), which makes sense as pH is a measure of acidity on a logarithmic sale.

*Therefore it may suffice to just focus on **alcohol**, **volatile.acidity** and **pH** when modeling a relationship between quality and chemical attributes of given white wine.* We will dive deep into this hypothesis visually and numerically as we progress in this data exploration further.

Before we start exploring these correlations deeper, le us visualize how each of attributes relate with quality with a  smooth "lm" to see if there are other features we can dd to above current main feature list.

```{r echo=FALSE}
# We make indiv plots of correlation of quality with each variable and than
# arrange them to display in one grid.

plot_list <- list()
for(i in 2:(length(wine)-1)) {
  plot_list[[(i-1)]] <- ggplot(data=wine, 
                               aes_string(y=colnames(wine)[i], x="quality")) + 
                          geom_smooth(method=lm) + ylab(colnames(wine)[i])
}
do.call(grid.arrange, plot_list)

```

Besides the already known features such as alcohol, pH, volatile.acidity (and their strongly correlated features), we notice that sulphates seem to stand out as potential candidate for furher analysis.

Let us start exploring these correlations deeper.

### Alcohol  

Alcohol has strongest correlation with wine which seems to make sense as typically wine's perception does relate to alcohol content.

To understand how alcohol may be related to quality, we can ask few questions about how are alcohol values distributed across quality levels:

1. What is the average alcohol content for each quality?
2. Is there a threshold difference in alcohol content across quality levels?

Let us answer first question of average alcohol content for each quality level.

**Average alcohol for each quality level**
```{r echo=FALSE}
tapply(wine$alcohol, wine$quality, mean)
```

**Visual boxplot of alcohol for each quality level**
```{r echo=FALSE}
ggplot(data=wine, aes(y=alcohol, x=factor(quality))) + geom_boxplot() 
```

We can see that **alcohol** content does seems to increase as **quality** level increases for higher quality wines with score of 6 and above.

Let us visualize this deeper with a scatter plot.

**Scatter plot of alcohol versus quality**
```{r echo=FALSE}
ggplot(data=wine, aes(x=quality, y=alcohol)) + geom_jitter(alpha=0.25) +
  geom_smooth(method='lm') 
``` 

We can see a good linear regression line with balanced increasing slope indicating **strong linear correlation** between **alcohol** and **quality**. Higher quality wine tend to have higher mean alcohol content threshold above 11.

### Voaltile.Acidity  

Volatile.acidity has good correlation with wine which seems to make sense as typically wine's  taste does relate to acid amount. 

To understand how volatile.acidity may be related to quality, we can ask few questions about how are volatile.acidity values are distributed across quality levels:

1. What is the average volatile.acidity content for each quality?
2. Is there a threshold difference in volatile.acidity content across quality levels?

Let us answer first question of average volatile.acidity content for each quality level.

**Average volatile.acidity for each quality level**
```{r echo=FALSE}
tapply(wine$volatile.acidity, wine$quality, mean)
``` 

**Visual boxplot of volatile.acidity for each quality level**
```{r echo=FALSE}
ggplot(data=wine, aes(y=volatile.acidity, x=factor(quality))) + geom_boxplot() 
``` 

We can see that **voaltile.acidity** content seems to decrease as **quality** level increases except for very low or very high quality wines. Therefore volatile.acidity has negative influence on quality of wine.

Let us visualize this deeper with a scatter plot.

**Scatter plot of volatile.acidity versus quality**
```{r echo=FALSE}
ggplot(data=wine, aes(y=volatile.acidity, x=quality)) + geom_jitter(alpha=0.25) +
  geom_smooth(method='lm') 
``` 

We can see a decent linear regression line with decent decreasing slope indicating **good linear correlation** between **volatile.acidity** and **quality**. Therefore volatile.acidity has negative influence on quality of wines. Higher quality wine tend to have lower mean volatile.acidity content threshold below 0.3. 

This finding that low amount of volatile acidity is good for wine is not surprising as it has connection with aroma of wine. Volatile acidity is an unstable acid formed by dissolving carbon dioxide in water. Generally seen as a fault in wine since it is quite disagreeable when excessive, although a tiny amount can enhance aromas in wine. Therefore good wine is likely to contain a small amount of acetic acid where as low quality wine may contain larger amount.

### pH  

Now let us look at our last attribute **pH**. We can recall from earlier observations that
**pH** is not so strongly correlated as previous two attributes. 

To understand how pH may be related to quality, we can ask few questions about how are pH values are distributed across quality levels:

1. What is the average pH content for each quality?
2. Is there a threshold difference in pH content across quality levels?

Let us answer first question of average pH content for each quality level.

**Average pH for each quality level**
```{r echo=FALSE}
tapply(wine$pH, wine$quality, mean)
``` 

**Visual boxplot of pH for each quality level**
```{r echo=FALSE}
ggplot(data=wine, aes(y=pH, x=factor(quality))) + geom_boxplot() 
``` 


We can see that **pH** content seems to increase as **quality** level increases except for very low quality wines.

Let us visualize this deeper with a scatter plot.

**Scatter plot of pH versus quality**
```{r echo=FALSE}
ggplot(data=wine, aes(y=pH, x=quality)) + geom_jitter(alpha=0.25) +
  geom_smooth(method='lm') 
``` 

We can see a average linear regression line with small increasing slope indicating **average linear correlation** between **pH** and **quality**. Higher quality wine tend to have higher mean pH content threshold above 3.2.

**So overall we see that higher quality wines tend to have higher alcohol and pH but lower level of volatile acid.**

### Sulphates  

Now let us look at our other potentail candidate attribute **sulphates**. 

To understand how sulphates may be related to quality, we can ask few questions about how are sulphates values are distributed across quality levels:

1. What is the average sulphates content for each quality?
2. Is there a threshold difference in sulphates content across quality levels?

Let us answer first question of average sulphates content for each quality level.

**Average pH for each quality level**
```{r echo=FALSE}
tapply(wine$sulphates, wine$quality, mean)
``` 

**Visual boxplot of pH for each quality level**
```{r echo=FALSE}
ggplot(data=wine, aes(y=sulphates, x=factor(quality))) + geom_boxplot() 
``` 

We can see that **sulphates** content seems to increase as **quality** level increases except for high quality wines.

Let us visualize this deeper with a scatter plot.

**Scatter plot of sulphates versus quality**
```{r echo=FALSE}
ggplot(data=wine, aes(y=sulphates, x=quality)) + geom_jitter(alpha=0.25) +
  geom_smooth(method='lm') 
``` 

We can see a below average linear regression line with very small increasing slope indicating **below average linear correlation** between **sulpahtes** and **quality**. This coupled with the observation that there is not much to differentiat ebetween level of sulphates between higher andlower quality wiens, this featuremay not be worth pursuing as influencing factor in determining quality of wines.

**So overall we see that higher quality wines tend to have higher alcohol and pH but lower level of volatile acid.**

## Short Questions

### Talk about some of the relationships you observed in this part of the investigation. How did the feature(s) of interest vary with other features in the dataset?
We can see that **quality** has strong correlation with **alcohol** (0.44), **density**(-0.31), **chlorides** (-0.21), **volatile.acidity** (-0.19), **toal.sulfur.dioxide** (-0.17) and bit weaker correlation with **fixed.acidity** (-0.11), **residual.sugar** (-0.10), **pH** (0.10).


### Did you observe any interesting relationships between the other features (not the main feature(s) of interest)?
We can see that there are some very strong correlation among other features such as **density** and **residual.sugar**. We also see that **density** , **total.sulphur.dioxide**, **chlorides** and **residual.sugar** all have good correlation with **alcohol**. Similarly **pH** and **fixed.acidity** also have strong correlation between them, which make sense as pH is measure of acidity on a logarithmic sale.

### What was the strongest relationship you found?
Strongest correlation is between **density** and **residual.sugar** (0.84).

---

# Multivariate Plots and Analysis

Let us continue to dive deeper into relationships we have discovered so far, mainly how **alcohol**, **volatile.acidity** and **pH** influence our quality levels. To explore these relations from another angle, let us group quality levels into three major categories: low level, average level and high level wines.

## Classifying wines as bad, average and good 

Let us classify our wines as following:

1. **Bad**: Quality score up to 4
2. **Average**: Quality score between 5 and 6
3. **Good**: Quality score 7 and above

**Add another variable quality.cut to repersent wine classification**
```{r echo=FALSE, Multivariate_Plots}
wine$quality.cut <- cut(wine$quality, breaks=c(0,4,6,10),
                        labels=c("Bad Wine: (0,4]", "Average Wine: (4, 6]",
                                 "Good Wine: (6, 10]"))
table(wine$quality.cut)
```

As we can see majority of wines are in **average** category.

Let us visualize the relationship between our chosen attributes and quality as sliced by wine quality classification.

```{r echo=FALSE}
ggplot(wine, aes(y=volatile.acidity, x=alcohol, color=quality)) +
  geom_jitter(alpha=0.5) + facet_wrap(~quality.cut)
```

We can see that majority of good wines (third facet) are in lower right quadrant of the graph, where alcohol content is higher and volatile.acidity is lower. Simliarly we can observe that bad wines (first facet) are in lower left corner and average wines (facet 2) are in lower midlle. This support our hypothesis that good wines tend to have higher alcohol content and moderate to low volatile acid. We can further observe that a large cluster of good wine in lower right has alcohol content greater than 11 and volatile.acidity less than 0.45.



Continuing our visual exploration of pH, alcohol influence on quality.
```{r echo=FALSE}
ggplot(wine, aes(y=pH, x=alcohol, color=quality)) + geom_jitter(alpha=0.5) +
  facet_wrap(~quality.cut)
```

We can see that majority of good wines (third facet) are in middle right side of the graph, where alcohol content is higher and pH is moderate. Simliarly we can observe that bad wines (first facet) are in lower middle left side and average wines (facet 2) are right in midlle of graph. This supports our hypothesis that good wines tend to have higher alcohol content and moderate pH. We can further observe that a large cluster of good wine in right half has alcohol content greater than 11 and pH between 3.0 to 3.45.

**Now let us try to bring all these together in a single scatter 3D visualization**
```{r echo=FALSE}
# Needs scatterplot3d package
# Colors get darker from front (low quality) to back (higher quality)
#
scatterplot3d(wine$alcohol, wine$quality.cut, wine$volatile.acidity, main="3D Scatterplot", highlight.3d=T, type="h", pch=16, angle=60)
```

We see that most of good wines (seen as black dots) are clustered in right lower back of 3d cube indicating that good wines tend to have higher alcohol and moderate to low volatile acid.

To sumamrize our findings so far, we are heading to a trend that indicates that majority of good wines have higer alcohol content is higher (greater than 10.75), volatile.acidity is lower (less than 0.5) and moderate pH (3 to 3.5).

## Model

Now that we have set of attributes which seem to have good correlation with quality. We can try to create a model using linear regression.

**quality = C0 + C1 * alcohol + C2 * volatile.acidity + C3 * pH**

```{r echo=FALSE}
m1 <- lm(formula = quality ~ alcohol, data = wine)
m2 <- lm(formula = quality ~ alcohol + volatile.acidity, data = wine)
m3 <- lm(formula = quality ~ alcohol + volatile.acidity + pH, data = wine)
mtable(m1, m2, m3)
```

We can see that we are getting **R^2** value of **0.24** with our 3rd model (**m3**) which models the above linear quality equation with three attributes (alcohol, volatile.acidity and pH). We can also see that attribute **pH* is not having significant impact as difference in R^2 value between 2nd model (m2) and 3rd model (m3) is quite small.

Let us examine 3rd model deeper.

```{r echo=FALSE}
summary(m3)
coef(m3)
```

We can see that our linear equation is:

**quality ~ 2.336 + 0.321 * alcohol - 1.965 * volatile.acidity + 0.223 * pH**
 
We also see that we have good degree of confidence in most of above coefficients. Large range of residual (min of -3.4 to max of 3.14) as compared to IQR of around 1 with median close to 0.04 seems to suggest some outliers may be spoiling our model. 

Let us look at our model visually.

```{r echo=FALSE}
par(mfrow=c(2,2)) # put two plots in a row
plot(m3)
```

We can see that first plot (residual v/s fitted) does indicate close to zero line on average with residuals spread out on both sides of this trend line. Similarly we see in second Q-Q graph that good normal distribution of residuals as most of them do line except bit on lower and higher side (outliers?). We also see small cook's distance other than few outliers. Therefore we can say we our linear model is a decent fit.

Let us visually see how each of attributes impacts the output (quality) in our model.

** Influence on predicted output quality (repersented as partial on y-axis)** 
```{r echo=FALSE}
termplot(m3)
```

We can see that while **alcohol** positively impacts **quality** but **volatile.acidity** negatively impacts **quality**. The **pH** has very small positive impact on **quality**. These results are not surprising and are very much in line with our correlation analysis earlier.

## Short Questions

### Talk about some of the relationships you observed in this part of the investigation. Were there features that strengthened each other in terms of looking at your feature(s) of interest?
**Alcohol** and **volatile.acidity** strengthened each other to impact the quality of wine.   

### Were there any interesting or surprising interactions between features?
Although not completely surprising **pH* did not influence *quality* as much. Also saw that there was a pattern combination of alcohol and volatile acidity when it came to determining quality of wines. Generally higher alcohol and moderate volatile.acidity led to higher quality of white wine.


### OPTIONAL: Did you create any models with your dataset? Discuss the strengths and limitations of your model.
Yes, created a linear regression model that uses the **alcohol**, **volatile.acidity** and **pH** of white wine to predict its quality. Here is the linear equation the model came up with:

**quality ~ 2.336 + 0.321 * alcohol - 1.965 * volatile.acidity + 0.223 * pH**

This model is not the strongest and can be improved further. Would like to see **R^2** of **0.24** go up further possibly with a polynomial regression (e.g. cubic equation), combination of other features not considered. Several outliers are playing spoilsport in prediction with this model. A larger data-set along with some other attributes which relate better with taste and body of white wine may help further to build a better predictive model.

---

# Final Plots and Summary

We have chosen to select three plots that lead us gradually to our final conclusion on how we can predict the quality of white wine based on given chemical properties. 


## Plot One

```{r echo=FALSE, Plot_One}
cor.plot(cor(wine[2:13]), numbers=T,
         main="Correlation Matrix Plot for White Wine")
```

**DESCRIPTION:  This correlation matrix plot provides nice visual view with darker shaded color indicated stronger correlation.  We can see that *quality* has strong correlation with *alcohol* (0.44), *density*(-0.31), *chlorides* (-0.21), *volatile.acidity* (-0.19), *toal.sulfur.dioxide* (-0.17) and bit weaker correlation with *fixed.acidity* (-0.11), *residual.sugar* (-0.10), *pH* (0.10).**

We also see that **density** (-0.78), **total.sulphur.dioxide** (-0.45), **chlorides** (-0.36) and **residual.sugar** (-0.45) all have good correlation with our strongly correlated attributed **alcohol**. Similarly **pH** has strong correlation (-0.43) with **fixed.acidity**. Therefore it may suffice to just focus on **alcohol**, **volatile.acidity** and to lesser extent **pH** as set of attributes which have most independent influence on quality of white wine.

## Plot Two

```{r echo=FALSE, Plot_Two}
g1 <- ggplot(data=wine, aes(y=quality, x=alcohol)) + 
      geom_jitter(alpha=0.25, color='red') + 
      geom_smooth(method='lm') + 
      ylab('Quality (0:very bad to 10:very excellent)') +
      xlab('Alcohol (% volume)')
g2 <- ggplot(data=wine, aes(x=volatile.acidity, y=quality)) + 
      geom_jitter(alpha=0.25, color='magenta') + geom_smooth(method='lm') +
      ylab('Quality (0:very bad to 10:very excellent)') +
      xlab('Volatile Acidity (g / dm^3)')
grid.arrange(g1, g2, nrow=1, 
             main="Impact of Alcohol, Volatile Acidity on White Wine Quality")

```

**DESCRIPTION: We can see a good linear regression lines with increasing slope for alcohol and decreasing slope for volatile acidity indicating strong linear correlation of these features with quality of white wine. Higher quality wine tend to have higher  alcohol content and lower volatile acidity**

That alcohol has strongest correlation with wine seems to make sense as typically wine's perception does relate to alcohol content. Simliarly the finding that low amount of volatile acidity is good for wine is not surprising as it has connection with aroma of wine. Volatile acidity is an unstable acid formed by dissolving carbon dioxide in water. Generally seen as a fault in wine since it is quite disagreeable when excessive, although a tiny amount can enhance aromas in wine. Therefore good wine is likely to contain a small amount of acetic acid where as low quality wine may contain larger amount.


## Plot Three
 
```{r echo=FALSE, Plot_Three}
# We are excluding top and bottom 0.5% observations as outliers here
# x and y intercept line values are decided based on our earlier observations
# during multivariate analysis
#
ggplot(data=wine, aes(x=alcohol, y=volatile.acidity, color=quality.cut)) +
  geom_jitter(alpha=0.5) + coord_cartesian( 
    xlim=c(quantile(wine$alcohol,.005), quantile(wine$alcohol,.995)),
    ylim=c(quantile(wine$volatile.acidity,.01), 
           quantile(wine$volatile.acidity,.99)) ) +
  geom_vline(xintercept = 10.75, linetype='dashed', color='brown', alpha=0.5) +
  geom_hline(yintercept = 0.5, linetype='dashed', color='brown', alpha=0.5) + 
  ylab('Volatile Acidity (g / dm^3)') +
  xlab('Alcohol (% volume)') +
  ggtitle('Wine Quality: Volatile.Acidity v/s Alcohol')
```

**DESCRIPTION: This is a multivariate plot between *volatile.acidity* and *alcohol* whose points are colored appropriately with *quality* buckets. This plot provides a good summary visual view of how alcohol and volatile.acidity related to different quality buckets of white wine.**

We can summarize following from above plot:

1. **Good Wines (having score above 6)**: *These mostly lie in lower right quadrant (volatile.acidity less than 0.5 and alcohol content larger than 10.75). Therefore good white wine tends to have higher amount of alcohol and moderate to low amount of volatile acid.*

2. **Average Wines (having score from 5 to 6)**: *These mostly lie in lower left quadrant (volatile.acidity less than 0.5 and alcohol content less than 10.75). Therefore average white wine tends to have lower amount of alcohol and moderate to low amount of volatile acid.*

3. **Bad Wines (having score less than 5)**: **These mostly lie in upper left quadrant (volatile.acidity more than 0.5 and alcohol content less than 10.75). Therefore poor quality white wine tends to have lesser amount of alcohol and high amount of volatile acid.*

**So in summary trying to answer our question "*Which chemical properties influence the quality of white wines?*", we have found out that *alcohol* and *volatile.acidity* influence our *quality* of wine significantly. We also observed that a large cluster of good wines have alcohol content greater than 10.75 and volatile.acidity less than 0.5.**

---

# Reflections  

Determination of wine quality is quite a complex inter-play of culture, society, reputation of a region or vineyard along with its chemical components which give wine its distinctive taste and aroma. The whole chemical composition of a wine embodies the various stages of the wine producing process, including the grape variety, yeast strain, the containers used for fermentation and storage, and the ecological practice. Human sense of taste  and aroma is not so well understood. This means wine classification is typically left to wine experts.

While rating of a wine (quality) by experts is ultimately a subjective matter, but it was an interesting exercise to analyze the different chemical components and their inter-play in wine to come up with best guess on wine rating.  We were able to identify few patterns which can help int his regard. While we have been able to find a way to relate quality of white wines with alcohol and volatile acid content, there are certain outliers which indicate that there must be other factors in the play which determine white wine quality. These factors such as smell, taste, soil and climate conditions where grapes were grown, etc, which are not captured in this data-set, may influence the white wine quality.

Was happy to see that initial guess of alcohol, pH and various acids influencing the quality of wine was not too far off the mark. Interesting to see that while pH, which is a measure of acidity, did not influence quality as much as volatile acid did, may be because beyond certain quantity acetic acid spoiled the taste and aroma of white wine. This kind of relationship is difficult to capture in a linear model. Our linear model with R^2 of 0.24 was a good start but far from a strong model. Several outliers are playing spoilsport in prediction with this model. While with our linear model it is harder to predict exact quality of white wine with high accuracy, we had better success in finding a pattern to classify wine in broader terms such as good, average and poor. Our visual cluster analysis gave strong clues to a good wine having higher alcohol content and moderate to low volatile acid.

This model is not the strongest and can be improved further. Would have liked to see much higher R^2. A better pruning of outliers can help in making our model better. Would like to explore other models such as polynomial regression (e.g. cubic equation) along with combination of other features not considered.  A larger data-set along with some other attributes which relate better with taste and body of white wine may help further to build a better predictive model. Would love to explore more ways to build a predictive model other than our linear regression model such as machine learning or SVM, etc. It will be interesting to see if our observations and model holds good for other kind of white wines outside the region this data-set came from. 

Would like to explore if we can use our model for fraud detection, crude wine classification and possibly enhance the training of students trying to become wine experts!

# References

Besides the Udacity course material in "Data Analysis with R", here is a list of some of the material explored while working on this project.

1. http://www.vinhoverde.pt/en/promotional-materials
2. http://www.cookbook-r.com/Graphs/index.html
3. http://www.physics.csbsju.edu/stats/box2.html
4. http://www.inside-r.org/packages/cran/psych/docs/00.psych
5. http://zief0002.github.io/Computing-Club/notes/2015_S/2015_S_02_Visual_Diagnostics.html


