---
title: "White Wine Exploratory Data Analysis"
author: "Lekhraj Sharma"
date: "Sept 2015"
output:
  html_document:
    keep_md: yes
    theme: cerulean
    toc: yes
---

```{r, echo=FALSE}
library(ggplot2)
library(psych)
```
## About the dataset  

Dataset chosen is "White Wine Quality" tidy data set downloaded from [Udacity site](https://s3.amazonaws.com/udacity-hosted-downloads/ud651/wineQualityWhites.csv). This publicly availale data set was created by Paulo Cortez, etc in 2009 for white variants of the Portuguese "Vinho Verde" wine. Due to privacy and logistic issues, only physicochemical (inputs) and sensory (the output) variables are available (e.g. there is no data about grape types, wine brand, wine selling price, etc.).

This tidy data set contains 4,898 white wines with 11 variables on quantifying the chemical properties of each wine. At least 3 wine experts rated the quality of each wine, providing a rating between 0 (very bad) and 10 (very excellent).

More information about this data set is available at [Udacity site](https://s3.amazonaws.com/udacity-hosted-downloads/ud651/wineQualityInfo.txt)


### Variable description  

Description of attributes:  

1. **Fixed acidity** (tartaric acid - g / dm^3) : most acids involved with wine or fixed or nonvolatile (do not evaporate readily)  
2. **Volatile acidity** (acetic acid - g / dm^3) : the amount of acetic acid in wine, which at too high of levels can lead to an unpleasant, vinegar taste  
3. **Citric acid** (g / dm^3) : found in small quantities, citric acid can add 'freshness' and flavor to wines  
4. **Residual sugar** (g / dm^3) : the amount of sugar remaining after fermentation stops, it's rare to find wines with less than 1 gram/liter and wines with greater than 45 grams/liter are considered sweet  
5. **Chlorides** (sodium chloride - g / dm^3): the amount of salt in the wine  
6. **Free sulfur dioxide** (mg / dm^3): the free form of SO2 exists in equilibrium between molecular SO2 (as a dissolved gas) and bisulfite ion; it prevents microbial growth and the oxidation of wine  
7. **Total sulfur dioxide** (mg / dm^3): amount of free and bound forms of S02; in low concentrations, SO2 is mostly undetectable in wine, but at free SO2 concentrations over 50 ppm, SO2 becomes evident in the nose and taste of wine  
8. **Density** (g / cm^3): the density of water is close to that of water depending on the percent alcohol and sugar content 
9. **pH**: describes how acidic or basic a wine is on a scale from 0 (very acidic) to 14 (very basic); most wines are between 3-4 on the pH scale  
10. **Sulphates** (potassium sulphate - g / dm3): a wine additive which can contribute to sulfur dioxide gas (S02) levels, wich acts as an antimicrobial and antioxidant  
11. **Alcohol** (% by volume): The percent alcohol content of the wine  
12. **Quality** (score between 0 to 10) : output variable based on sensory data, rating between 0 (very bad) and 10 (very excellent)
   
## Questions  

By performing this analysis, we seek to answer the following questions:  

1. *Which chemical properties influence the quality of white wines?*

## Data exploration  

Let us begin by loading the dataset for white wines in a data frame.


```{r}
wine <- read.csv("./wineQualityWhites.csv")
```

### Wine quality    

Let us start with getting basic descriptive statistics about the dataset. We will pay extra attention to **quality** as we are trying to answer question on how can we determine wine quality.


```{r}
summary(wine)
describe(wine[2:12], skew=F)
lowerCor(wine)
pairs.panels(wine)
```

Both `summary` and `describe` commands give us interesting insights about our attributes. Clearly attribute **X** is just a serial id for each observation. All other attributes except **quality** have a continous value with their mean and median being close, therefore the values are distributed evenly across the total range. **Quality* has discrete values ranging from 3 to 9. 

Looking further in pairwise correlation via `lowerCor`, we can see that **quality** has strong correlation qith **alcohol** (0.44), **density**(-0.31), **chlorides** (-0.21), **volatile.acidity** (-0.19), **toal.sulfur.dioxide** (-0.17) and bit weaker correlation with **fixed.acidity** (-0.11), **residual.sugar** (-0.10), **pH** (0.10).

Our paiwise plot via `pairs.panel` gives us birdeye visual view into how each variable is ditributed along with how they relate to each other. We can see that almost all attributes other than **residual.sugar** and **alcohol** seem to have normal distribution. Both **residual.sugar** and **alcohol** are left skewed and can be converted to normal distribution with some transformations. 

We can see that there are some strong correlation among other attributes such as 0.84 between **density** and **residual.sugar**, which canbe seen in pairwise graph as well in terms almost 45 degree line between these two attributes. Since her ewe are focussing on **quality** relation with other attributes, we are **not** going to explore these relationships by themselves other than in context of attributes that relate to **quality**.

When we look further in pairwise graph and correaltion matrix, we can see that several of the strongly correlated attributes to **quality** have good correlation among themselves. Since we want to identify indepdent variables which impact the quality, it will be good to prune these out and select one with strongest correlation with **quality**. We see that **density** (-0.78), **total.sulphur.dioxide** (-0.45), **chlorides** (-0.36) and **residual.sugar** (-0.45) all have good correlation with our strongly correlated attributed **alcohol**. Simliarly **pH** and **fixed.acidity** also have strong correlation (-0.43), which make sense as pH is measure of acidity on a logarithmic sale. *Therefore it may suffice to just focus on **alcohol**, **volatile.acidity** and **pH** when modeling a relationship between quaity and chemical attributes of given white wine. We will dive deep into this hypothesis visually and numerically as we progress in this data exploration further.*

Let us first looking into distribution of quality deeper.

```{r}
summary(wine$quality)
IQR(wine$quality)
table(wine$quality)
```

So **quality** discrete values range between 3 and 9 with no scores below 3 or above 9. Since both max and min are more than 1.5*IQR away from their respective quartile, it seems we do have outliers here. From `table` we can see we have around 4% (200/4898) outliers which is significant. 

Let us plot all this to visualize this better. 

```{r}
ggplot(data=wine, aes(x=quality)) + geom_bar(binwidth=1, color='black', fill='green') +
  geom_vline(xintercept = mean(wine$quality), linetype= 'longdash', color='red', alpha=.5)+               
  geom_vline(xintercept = median(wine$quality), linetype='longdash', color='blue', alpha=.5) + 
  geom_vline(xintercept = quantile(wine$quality, 0.25) - 1.5 * IQR(wine$quality),
             linetype='dashed', alpha=.5) +
  geom_vline(xintercept = quantile(wine$quality, 0.75) + 1.5 * IQR(wine$quality), 
             linetype='dashed', alpha=.5)
```

We can see that majority of wines have quality scores ranging from 5 to 7. Therefore there are lot more average quality wines and few poor or excellent quality wines. 

### Influential variables  

Let have a look at how correlated our different variables are.  

```{r}
cor(x=reds[,2:12], y=reds$quality)
```

From previous we see that the following variables are correlated with
quality:  

- Alcohol (+++)  
- Volatile acidity (---)  

- Citric acid (++)

- Fixed acidity (+)  
- Sulphates ?? worht investigating (+)  
- Total sulphur dioxide (-)  
- Density (-)  
- Chlorides (-)  

We will concentrate in the first three variables that show stronger correlation
with quality.  


#### Alcohol  

Alcohol contributes to a wine's *body* perception during taste and also to its 
balance vs acidity and tanicity (just for reds).  

First of all, a quick summary of the differnet alcohol values.  

```{r}
summary(reds$alcohol)
reds_alcohol_mean <- mean(reds$alcohol)
reds_alcohol_median <- median(reds$alcohol)
```

In order to understand the relationship between alcohol and quality, we will do
two things. First we will inspect how the alcohol values are distributed accross 
the different quality scores. Then we will analyse how they change together.  

For example, what is the average alcohol level for each quality score? We can 
quickly see this by using the `dplyr` library or just by using the `tapply` command.  

```{r}
tapply(reds$alcohol, reds$quality, mean)
```

And in a visual way, alcohol by quality level, together with the median for the 
whole distribution and the mean for the quality scores.  

```{r}
ggplot(data=reds, aes(x=quality_as_factor, y=alcohol)) +
    geom_boxplot() +
    geom_hline(show_guide=T, yintercept=reds_alcohol_mean, linetype='longdash', alpha=.5, color='blue') +
    geom_vline(xintercept = reds_quality_mean-reds_quality_min+1, linetype='longdash', color='blue', alpha=.5) +
    xlab("Wine Quality") +
    ylab("Alcohol")
```

We can see how wine scores that are beyond the mean quality value of 
`r reds_quality_mean`, also tend to show values beyond the mean alcohol 
value of `r reds_alcohol_mean`.  

We already see a tendency in the boxplots. But let's use a scatter plot here, 
including a linear regression line.   

```{r}
ggplot(data=reds, aes(x=as.numeric(quality), y=alcohol)) +
    geom_jitter(alpha=1/3) +
    geom_smooth(method='lm', aes(group = 1))+
    geom_hline(yintercept=reds_alcohol_mean, linetype='longdash', alpha=.5, color='blue') +
    geom_vline(xintercept = reds_quality_mean, linetype='longdash', color='blue', alpha=.5) +
    xlab("Wine Quality") +
    ylab("Alcohol")
```

#### Citric acid  

And as a third variable, let's perform a similar analysis for citric acid. Remember
that this one is not so strongly correlated as the previous ones.  

```{r}
summary(reds$citric.acid)
reds_citric.acid_mean <- mean(reds$citric.acid)
reds_citric.acid_median <- median(reds$citric.acid)
reds_citric.acid_sd <- sd(reds$citric.acid)
reds_citric.acid_max <- max(reds$citric.acid)
caout <- (reds_citric.acid_max - reds_citric.acid_mean) / reds_citric.acid_sd
```

The mean value for the citric acid is `r reds_citric.acid_mean` with
a standard deviation of `r reds_citric.acid_sd`. As a note, the maximum value 
for citric acid is `r reds_citric.acid_max`. That is `r caout` standard
deviations beyond the mean! Again, probably an outlier.   

And now the relationship between this variable and the quality score.  

```{r}
tapply(reds$citric.acid, reds$quality, mean)
```

And in a visual way, citric acid by quality level, together with the mean for the 
whole distribution and the mean for the quality scores.  

```{r}
ggplot(data=reds, aes(x=quality_as_factor, y=citric.acid)) +
    geom_boxplot() +
    geom_hline(show_guide=T, yintercept=reds_citric.acid_mean, linetype='longdash', alpha=.5, color='blue') +
    geom_vline(xintercept = reds_quality_mean-reds_quality_min+1, linetype='longdash', color='blue', alpha=.5) +
    xlab("Wine Quality") +
    ylab("Citric Acid")
```

```{r}
ggplot(data=reds, aes(x=as.numeric(quality), y=citric.acid)) +
    geom_jitter(alpha=1/3) +
    geom_smooth(method='lm', aes(group = 1))+
    geom_hline(yintercept=reds_citric.acid_mean, linetype='longdash', alpha=.5, color='blue') +
    geom_vline(xintercept = reds_quality_mean, linetype='longdash', color='blue', alpha=.5) +
    xlab("Wine Quality") +
    ylab("Citric Acid")
```

Again, we can see a tendency. Good wines tend to have higher citric acid levels.  

### Faulty wines  

What characteristics can we see influence wine quality negatively?  

#### Voltile acidity  

Let's perform a similar analysis for volatile acidity.  

```{r}
summary(reds$volatile.acidity)
reds_volatile.acidity_mean <- mean(reds$volatile.acidity)
reds_volatile.acidity_median <- median(reds$volatile.acidity)
reds_volatile.acidity_sd <- sd(reds$volatile.acidity)
reds_volatile.acidity_max <- max(reds$volatile.acidity)
vaout <- (reds_volatile.acidity_max - reds_volatile.acidity_mean) / reds_volatile.acidity_sd
```

The mean value for the volatile acidity is `r reds_volatile.acidity_mean` with
a standard deviation of `r reds_volatile.acidity_sd`. As a note, the maximum value 
for volatile acidity is `r reds_volatile.acidity_max`. That is `r vaout` standard
deviations beyond the mean! Probably an outlier...  

And now the relationship between this variable and the quality score.  

```{r}
tapply(reds$volatile.acidity, reds$quality, mean)
```

And in a visual way, volatile acidity by quality level, together with the median for the 
whole distribution and the mean for the quality scores.  

```{r}
ggplot(data=reds, aes(x=quality_as_factor, y=volatile.acidity)) +
    geom_boxplot() +
    geom_hline(show_guide=T, yintercept=reds_volatile.acidity_mean, linetype='longdash', alpha=.5, color='blue') +
    geom_vline(xintercept = reds_quality_mean-reds_quality_min+1, linetype='longdash', color='blue', alpha=.5) +
    xlab("Wine Quality") +
    ylab("Volatile Acidity")
```

```{r}
ggplot(data=reds, aes(x=as.numeric(quality), y=volatile.acidity)) +
    geom_jitter(alpha=1/3) +
    geom_smooth(method='lm', aes(group = 1))+
    geom_hline(yintercept=reds_volatile.acidity_mean, linetype='longdash', alpha=.5, color='blue') +
    geom_vline(xintercept = reds_quality_mean, linetype='longdash', color='blue', alpha=.5) +
    xlab("Wine Quality") +
    ylab("Volatile Acidity")
```

In each step we can see the negative influence of volatile acidity in a 
wine's quality score.  

#### Another way to look at it  

Let us say we consider faulty wines those that do not get a score of at least 5
in the scale given (0 to 10). Let us then define a new variable faulty that takes
a boolean value based on that.  

```{r}
reds <- transform(reds, is.faulty = quality<5)
```

How many faulty wines do we have?  

```{r}
faulty.reds <- table(reds$is.faulty)
```

We can see a faulty wines ratio of `r faulty.reds[2]/faulty.reds[1]` .  

From those variables we determined were highly correlated, let us see which ones
are characteristic of faulty wines and at what levels. Boxplots are ideal for 
this purpose. Let us start with *volatile acidity*.  

```{r}
ggplot(data=reds, aes(x=is.faulty, y=volatile.acidity)) +
    geom_boxplot() +
    geom_hline(show_guide=T, yintercept=reds_volatile.acidity_mean, linetype='longdash', alpha=.5, color='blue') +
    geom_vline(xintercept = reds_quality_mean-reds_quality_min+1, linetype='longdash', color='blue', alpha=.5) +
    xlab("Faulty Wine") +
    ylab("Volatile Acidity")
summary(subset(reds, is.faulty)$volatile.acidity)
```


What about *alcohol*? Maybe there is a minimum threshold under that a wine is perceived as faulty?   

```{r}
ggplot(data=reds, aes(x=is.faulty, y=alcohol)) +
    geom_boxplot() +
    geom_hline(show_guide=T, yintercept=reds_alcohol_mean, linetype='longdash', alpha=.5, color='blue') +
    geom_vline(xintercept = reds_quality_mean-reds_quality_min+1, linetype='longdash', color='blue', alpha=.5) +
    xlab("Faulty Wine") +
    ylab("Alcohol")
summary(subset(reds, is.faulty)$alcohol)
```

In this case the answer is clearly no. Although faulty wines tend to have slightly
less alcohol than non faulty ones, their ranges pretty much overlap. We will talk a
bit more about this in the summary. 

These were the two mainly correlated variables. But let us explore some other variables.  

Next let us try with *citric acid* that typically defines fresh wines.  

```{r}
ggplot(data=reds, aes(x=is.faulty, y=citric.acid)) +
    geom_boxplot() +
    geom_hline(show_guide=T, yintercept=reds_citric.acid_mean, linetype='longdash', alpha=.5, color='blue') +
    geom_vline(xintercept = reds_quality_mean-reds_quality_min+1, linetype='longdash', color='blue', alpha=.5) +
    xlab("Faulty Wine") +
    ylab("Citric Acid")
summary(subset(reds, is.faulty)$citric.acid)
```

This might be and important one, yes. We can confirm this using the correlation
between both variables.  

```{r}
cor(x=reds$volatile.acidity, y=reds$citric.acid)
```  


## Summary

Recalling our original questions:  

1. How is the quality of the wines tasted?  
2. What is the minimum set of properties and their values that defines a high quality wine?  
3. What are considered wine defects?  
Citric acid seems to influence wine quality in a way that, when is too low, the wine
is considered faulty.  

We already saw that our dataset contains wines over average. This is specially clear
in the scores distribution chart.  

```{r}
library(ggplot2)
ggplot(data=reds, aes(x=quality)) +
    geom_bar(binwidth=1, color='black', fill='white') +
    coord_cartesian(xlim=c(0,10)) +
    geom_vline(xintercept = reds_quality_median, linetype='longdash', alpha=.5) +
    geom_vline(xintercept = reds_quality_q1 - 1.5*reds_quality_iqr, linetype='longdash', alpha=.5) +
    geom_vline(xintercept = reds_quality_q3 + 1.5*reds_quality_iqr, linetype='longdash', alpha=.5) +
    geom_vline(xintercept = reds_quality_mean, linetype=1, color='blue', alpha=.5) +
    xlab("Wine Quality") +
    ylab("# Samples")
```

So if we consider the average quality score to be 5 in a 0 to 10 scale, we can 
see that our quality scores are beyond average under multiple criteria. 

About good wines. It seems clear that they have certain levels of alcohol and citric
acid, and lack of defects. We already saw charts and distribution in previous 
sections showing a positive correlation between these variable and quality. 
So let's try a new plot here. Let's plot alcohol vs citric acid so we devide the
chart area y quadrants. Then let's color each plot using the quality score and see
where good wines fall.  

```{r}
reds$quality.cut <- cut(reds$quality, breaks=c(0,4,6,10))
ggplot(data=reds, aes(x=alcohol, y=citric.acid)) +
    coord_cartesian(
        xlim=c(quantile(reds$alcohol,.01),quantile(reds$alcohol,.99)),
        ylim=c(quantile(reds$citric.acid,.01),quantile(reds$citric.acid,.99))
        ) +
    geom_jitter(alpha=.5, aes(size=quality.cut, color=quality.cut)) +
    geom_vline(xintercept = reds_alcohol_mean, linetype='longdash', color='blue', alpha=.5) +
    geom_hline(yintercept = reds_citric.acid_mean, linetype='longdash', color='blue', alpha=.5) +
    xlab("Alcohol") +
    ylab("Citric Acid")
```

In order to improve clarity, we have defined a cut in the scores.  

Something interesting emerge from this chart. And is that
good wines concentrate in the ipper right quadrant. That is, even when alcohol and
citric acid sometimes produce high quality wines in isolation it is when we have
certain levels of both when we have a more high quality scores. This is related 
with the concept of gustative **balance** in a wine, where the alcohol body and 
sweet feeling balances with citric acid.  

What about defects? We already saw thet *volatile acidity* is very determinant
for faulty wines. We also saw that alcohol is not. Both faulty and correct wines
have similar distributions of levels of alcohol. And we finally saw that a high
level of citric acid is associated with less faulty wines.  

What about wines that, under the previous premises, have
high level of volatile acidity? Our plot is getting a bit too complicated, so
let's try using tables. We will make use of our previously defined 
variable `isFaulty`. We also want to assign one of the previous quadrants to each
wine. They are separated by the axis mean values, so it should be easy.    

```{r}
reds <- transform(
        reds,
        quadrant = ifelse(alcohol>reds_alcohol_mean,
                          ifelse(citric.acid>reds_citric.acid_mean,4,2),
                          ifelse(citric.acid>reds_citric.acid_mean,3,1))
    )
```

These are the quadrants:  

  * Quadrant 1: wines low in alcohol and citric acid.  
  * Quadrant 2: wines high in alcohol but low in citric acid.  
  * Quadrant 3: wines low in alcohol but high in citric acid.  
  * Quadrant 4: wines in balance with high alcohol and citric acid.  
  
Let's table each quadrant to see what we have.  

```{r}
table(reds$quadrant)
```

Let's see the proportions of faulty wines in each quadrant.  

```{r}
sort(tapply(reds$is.faulty, reds$quadrant, mean))
```

That is, there is a lower proportion of faulty wines in the 4th quadrant, etc.  

So finally, what about the volatile acidity level?

```{r}
sort(tapply(reds$volatile.acidity>reds_volatile.acidity_mean, reds$quadrant, mean))
```

Again, the proportion of wines with a high level of volatile acidity, typically produced
by acetic bacteries present in vinegar, is lower in the 4th quadrant and higher in 
the 1st one. We can also see that there is a considerable gap between 3rd quadrant
and the lower two, due to the reduction effect of citric acid on volatile 
acidity (negative correlation).  

```{r}
cor(x=reds$volatile.acidity, y=reds$citric.acid)
```

## Reflections  

There are many other factors that are related with good wines. Many of them are related with
smells and flavours and not with chemical properties and gustative perceptions like 
these that we have in our dataset. Although our variables are kind of explanatory of
what we have, we have also seen some cases where the must be other explanations for high
or low quality levels.  

However, within our limitations, we have discover the very interesting concept
of gustative balance between alcohol and acidity, and how important is for a wine
to be free of defects.  


